Supplementary Materials

### 2023-02-17 analysis/paper/_results.qmd
prepare preprint

auth_plot <- uhplc_data_long %>%
  filter(
    batch == "batch2",
    !compound %in% c("cbd", "cbn", "cocaine", "thc", "thca-a", "thcva") # remove compounds not detected in batch 2
    ) %>% 
  semi_join(quant_filter, by = c("sample", "compound")) %>% # remove compounds not detected in each sample 
  mutate(
    extraction = factor(extraction, levels = c("wash1", "wash2", "wash3", "calc")),
    #sample = as.factor(sample)
    ) %>% 
  ggplot(aes(x = extraction, y = quant, group = sample, colour = sample)) +
    geom_line() +
    geom_point(size = 0.2) +
    #facet_wrap(~ compound, scales = "free_y", labeller = compound_name_repair) +
    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names), dir = "v") +
    theme_bw() +
    scale_colour_viridis_d() +
    scale_y_continuous(position = "right") +
    labs(y = "Quantity (ng)") +
    theme(
      legend.position = "none",
      axis.title.x = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
    )

compound_detect_plot <- uhplc_calculus_long %>%
  #remove_missing() %>%
  filter(quant > 0) %>% 
  ggplot(aes(x = compound, fill = batch)) +
    #geom_jitter(width = 0.2) +
    #geom_boxplot() +
    geom_bar(position = "dodge") +
    #facet_wrap(~ batch) +
    theme_bw() +
    scale_fill_viridis_d() +
    scale_x_discrete(labels = compound_names) +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
      legend.position = "left",
      panel.grid.major.x = element_blank()
      ) +
    labs(y = "Count")

@@

weight_quant_plot <- uhplc_calculus_long %>%
  filter(
    batch == "batch2",
    quant > 0,
    id != "MB329" # unusually high nicotine and cotinine quantity in batch 1
    ) %>% 
  ggplot(aes(x = weight, y = quant)) +
    geom_point(col = viridisLite::viridis(1)) +
    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names), dir = "v", ncol = 1) +
    theme_bw() +
    theme(axis.title.y = element_blank()) +
    labs(y = "Quantity (ng)", x = "Weight (mg)")

>>>>

auth_plot <- uhplc_data_long %>%
  filter(
    batch == "batch2",
    !compound %in% c("cbd", "cbn", "cocaine", "thc", "thca-a", "thcva") # remove compounds not detected in batch 2
    ) %>% 
  semi_join(quant_filter, by = c("sample", "compound")) %>% # remove compounds not detected in each sample 
  mutate(
    extraction = factor(extraction, levels = c("wash1", "wash2", "wash3", "calc")),
    #sample = as.factor(sample)
    ) %>% 
  ggplot(aes(x = extraction, y = quant, group = sample, colour = sample)) +
    geom_line() +
    geom_point(size = 0.2) +
    #facet_wrap(~ compound, scales = "free_y", labeller = compound_name_repair) +
    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names), dir = "v") +
    theme_bw() +
    scale_colour_viridis_d() +
    scale_y_continuous(position = "right") +
    labs(y = "Quantity (ng)") +
    theme(
      legend.position = "none",
      axis.title.x = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.y = element_blank()
    )

compound_detect_plot <- uhplc_calculus_long %>%
  #remove_missing() %>%
  filter(quant > 0) %>% 
  ggplot(aes(x = compound, fill = batch)) +
    #geom_jitter(width = 0.2) +
    #geom_boxplot() +
    geom_bar(position = "dodge") +
    #facet_wrap(~ batch) +
    theme_bw() +
    scale_fill_viridis_d() +
    scale_x_discrete(labels = compound_names) +
    scale_fill_viridis_d(labels = c(1,2)) +
    theme(
      axis.title.x = element_blank(),
      axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
      legend.position = "left",
      panel.grid.major.x = element_blank()
      ) +
    labs(y = "Count", fill = "Batch")

@@

weight_quant_plot <- uhplc_calculus_long %>%
  filter(
    batch == "batch2",
    quant > 0,
    id != "MB329" # unusually high nicotine and cotinine quantity in batch 1
    ) %>% 
  ggplot(aes(x = weight, y = quant)) +
    geom_point(col = viridisLite::viridis(1)) +
    facet_wrap(~ compound, scales = "free_y", labeller = labeller(compound = compound_names), dir = "v", ncol = 1) +
    theme_bw() +
    theme(
      axis.title.y = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank()
    ) +
    labs(y = "Quantity (ng)", x = "Weight (mg)")

### 2023-02-17 analysis/paper/paper.qmd
prepare preprint

@@

All raw data, analysis scripts, and the source code for the manuscript and
supplementary materials are available as a research compendium<!--Zenodo link-->
using the structure defined in the **rrtools** R package [@Rrrtools].

>>>>

This research has received funding from the European Research Council under the
European Union’s Horizon 2020 research and innovation program, grant agreement
number STG–677576 (“HARVEST”).

@@

All raw data is available on Zenodo (<https://doi.org/10.5281/zenodo.7648757>).
Analysis scripts, and the source code for the manuscript and
supplementary materials are available as a research compendium<!--Zenodo link-->
using the structure recommended by the **rrtools** R package [@Rrrtools].

### 2023-02-17 analysis/supplementary-materials/supp-mat.qmd
prepare preprint

---
title: "Supplementary Materials"
execute: 
  echo: false
  warning: false
knitr: 
  opts_chunk: 
    message: false
format: 
  html:
    toc: true
    code-fold: true
---

@@

### metadata

@@

| variable | description |
|---|---|
| id | unique identifier for the individual |
| sample | sample number for the UHPLC-MS/MS analysis |
| element | which tooth (FDI notation) was sampled for UHPLC-MS/MS analysis |
| replicated | Whether the sample was included in the replication batch (TRUE/FALSE) |
| batch1_weight | weight (mg) of sample in batch 1 |
| batch2_weight | weight (mg) of sample in batch 2 |

### dental-inv

@@

- p = present
- m = missing (for unknown reason)
- aml = ante-mortem loss
- dna = removed for DNA sampling

### caries

@@

- none = no caries present
- mes = mesial
- dis = distal
- lin = lingual (and labial)
- buc = buccal
- occ = occlusal
- crown = crown (caries-related loss of much of/entire tooth crown)
- root = root
- blank = tooth not present

### periodontitis

@@

### periapical

@@

### calculus_full

@@

Scored according to Standards.

### path-conditions

@@

```{r}
#| label: tbl-sampling
#| tbl-cap: "Table showing which individuals were sampled in this study (element) and which individuals were sampled in the previous study (KZ_element)."
metadata %>% 
  select(id, element, KZ_element) %>% 
  mutate(
    element = case_when(is.na(element) ~ FALSE,
                   TRUE ~ TRUE),
    KZ_element = case_when(is.na(KZ_element) ~ FALSE,
                   TRUE ~ TRUE),
    )
```

@@

The sample consists of `r nrow(demography)` individuals, most of which are
middle adult male individuals (@fig-age-distribution).

```{r}
#| label: fig-age-distribution
demography %>%
  ggplot(aes(x = age, fill = sex)) +
    geom_bar() +
    theme_minimal()
```

@@

Missing data across all dental datasets

```{r}
#| label: fig-missing-dental
#| fig-cap: "Number of missing values per individual and tooth across all dental scores (caries, periodontitis, periapical lesions, calculus)."
dental_long %>%
  mutate(across(c(caries, periodont, periap, calculus), is.na)) %>%
  select(c(caries, periodont, periap, calculus, id, tooth)) %>%
  #group_by(id, tooth) %>% 
  rowwise() %>% 
  mutate(missing = sum(caries, periodont, periap, calculus)) %>% 
  ggplot(aes(x = tooth, y = id, fill = missing)) +
    geom_tile()
```

@@

```{r}
#| label: tbl-lloq
#| tbl-cap: "Target compounds and lower limits of quantitation."
lloq %>%
  arrange(compound)
```

```{r}
#| label: tbl-uhplc-batch-1
#| tbl-cap: "Results from the UHPLC analysis first batch. Quantity of compound in the dissolved calculus, represented in ng."
uhplc_calculus_long %>%
  filter(batch == "batch1") %>%
  #select(sample, contains("batch1")) %>%
  #rename_with(~ str_remove(.x, "_calc_batch1"), .cols = contains("_calc_batch1"))
  pivot_wider(names_from = "compound", values_from = "quant") %>%
  select(!c(extraction, batch))
```

@@

```{r}
#| label: fig-corr-notches
# correlation between nicotine and cotinine conc, and pipe notches
uhplc_calculus_long %>%
  ggplot(aes(x = pipe_notch, y = conc)) +
    geom_point() +
    facet_wrap(~ compound + batch, scales = "free_y")
```

@@

# ratio of nicotine to cotinine
tobacco_ratio <- tobacco %>%
  select(id, batch, compound, conc) %>% 
  pivot_wider(names_from = compound, values_from = conc) %>% 
  group_by(id, batch) %>% 
  summarise(ratio = cotinine / nicotine) %>%
  remove_missing() %>% 
  left_join(select(demography, id, age, sex, preservation, pipe_notch)) %>% 
  filter(batch == "batch2",
         ratio != Inf) %>% 
  mutate(
    age = case_when(
      age == "eya" ~ 0,
      age == "lya" ~ 1,
      age == "ma" ~ 2,
      age == "old" ~ 3
    ),
    preservation = case_when(
      preservation == "fair" ~ 0,
      preservation == "good" ~ 1,
      preservation == "excellent" ~ 2
    )
  )

@@

In the replicated samples only, tobacco detection was successful in
`r nrow(filter(replicated_accuracy, pipe_notch == "Y", detection > 0))`
out of
`r nrow(filter(replicated_accuracy, pipe_notch == "Y"))`
pipe smokers
(`r scales::percent(nrow(filter(replicated_accuracy, pipe_notch == "Y", detection > 0)) / nrow(filter(replicated_accuracy, pipe_notch == "Y")), accuracy = 0.1)`)
Including individuals with absence of a pipe notch and concurrent absence of
compounds as a correct identification, gives an overall accuracy of
`r scales::percent(mean(replicated_accuracy$correct, na.rm = T), accuracy = 0.1)`.

@@

```{r}
#| label: fig-dental-plots
#| layout-ncol: 2
#| fig-cap: 
#|   - "Antemortem tooth loss in sample."
#|   - "Periodontal scores in the sample."
dental_long %>%
  dental_plot(fill = status)

@@

Brief statement about amtl at site.

```{r}
dental_long %>% 
  group_by(id) %>% 
  amtl_ratio(.status = status, .add = T)
```

@@

<!--add a caries correction?-->
Caries were scored as the location on each individual tooth. Multiple locations on
a single tooth were separated with `;`. The size of caries was also
recorded, but not used in further analysis. Large caries that cover multiple
surfaces with an unknown origin were recorded as 'crown'. Caries counts were
extracted based on the number of locations recorded.

@@

```{r}
caries_count %>%
  caries_ratio(.caries = count, class, id) %>%
  ggplot(aes(x = class, y = ratio)) +
    geom_violin(aes(fill = class)) +
    geom_boxplot(width = 0.1)
```

```{r}
upper_order <- c(paste0("t", 18:11), paste0("t", 21:28))
lower_order <- c(paste0("t", 48:41), paste0("t", 31:38))

@@

```{r}
#| label: fig-periodont
dental_long %>%
  filter(!is.na(periodont)) %>% 
  dental_plot(fill = periodont)
```

@@

>>>>

---
title: "Supplementary Materials"
execute: 
  echo: false
  warning: false
knitr: 
  opts_chunk: 
    message: false
format: 
  html:
    toc: true
    code-fold: true
  pdf:
    toc: true
bibliography: "../paper/references.bib"
---

@@

All raw data are available for download from Zenodo.

### metadata.csv

@@

| variable | description |
|---|---|
| id | unique identifier for the individual |
| sample | sample number for the UHPLC-MS/MS analysis |
| element | which tooth (FDI notation) was sampled for UHPLC-MS/MS analysis |
| KZ_element | which tooth (FDI notation) was sampled for aDNA in original study |
| replicated | Whether the sample was included in the replication batch (TRUE/FALSE) |
| batch1_weight | weight (mg) of sample in batch 1 |
| batch2_weight | weight (mg) of sample in batch 2 |

### dental-inv.csv

@@

- p = present
- m = missing (for unknown reason - likely postmortem loss)
- aml = ante-mortem loss
- dna = removed for DNA sampling

### caries.csv

@@

- none = no caries present
- mes = mesial
- dis = distal
- lin = lingual
- buc = buccal (and labial)
- occ = occlusal
- crown = crown (large caries lesion covering multiple surfaces)
- root = root
- blank = tooth not present

### periodontitis.csv

@@

### periapical.csv

@@

### calculus_full.csv

@@

### path-conditions.csv

@@

```{r}
#| label: tbl-sampling
#| tbl-cap: "Table showing which individuals were sampled in this study and which individuals were sampled in the previous study. When both TRUE, the individual was sampled twice."
metadata %>% 
  select(id, element, KZ_element) %>% 
  mutate(
    element = case_when(is.na(element) ~ FALSE,
                   TRUE ~ TRUE),
    KZ_element = case_when(is.na(KZ_element) ~ FALSE,
                   TRUE ~ TRUE),
    ) %>%
  knitr::kable(col.names = c("ID", "this study", "previous study"))
```

@@

The sample consists of `r nrow(demography)` individuals, most of which are
middle adult male individuals (@fig-age-distribution). Middle adult males were
preferentially targeted and the sample age and sex distribution is therefore not
representative of the population. This was done to limit potential confounding
factors, and because pipe notches, which served as a positive control for tobacco,
are predominantly seen in male individuals at the site.

```{r}
#| label: fig-age-distribution
#| fig-cap: "Distribution of age and sex in the sample. f = female; pf = probable female; pm = probable male; m = male; eya = early young adult (18-24 years); lya = late young adult (25-34 years); ma = middle adult (35-49 years); old = old adult (50+ years)."
demography %>%
  ggplot(aes(x = age, fill = sex)) +
    geom_bar() +
    theme_minimal()
```

@@

Missing data across all caries, periodontitis, and periapical datasets.

```{r}
#| label: fig-missing-dental
#| fig-cap: "Number of missing values per individual and tooth across all dental scores (caries, periodontitis, periapical lesions, calculus)."
dental_long %>%
  mutate(across(c(caries, periodont, periap), is.na)) %>%
  select(c(caries, periodont, periap, id, tooth)) %>%
  #group_by(id, tooth) %>% 
  rowwise() %>% 
  mutate(missing = sum(caries, periodont, periap)) %>% 
  ggplot(aes(x = tooth, y = id, fill = missing)) +
    geom_tile()
```

@@

```{r}
#| label: tbl-lloq
#| tbl-cap: "Target compounds and lower limits of quantitation (LLOQ)."
lloq %>%
  arrange(compound) %>% 
  knitr::kable(col.names = c("Target", "LLOQ"))
```

```{r}
#| label: tbl-uhplc-batch-1
#| tbl-cap: "Results from the UHPLC analysis first batch. Quantity of compound in the dissolved calculus, represented in ng."
uhplc_calculus_long %>%
  filter(batch == "batch1") %>%
  #select(sample, contains("batch1")) %>%
  #rename_with(~ str_remove(.x, "_calc_batch1"), .cols = contains("_calc_batch1"))
  select(!c(sample, conc, extraction, batch, presence)) %>%
  pivot_wider(names_from = "compound", values_from = "quant") %>% 
  knitr::kable()
```

@@

```{r}
#| label: fig-corr-notches
# correlation between nicotine and cotinine conc, and pipe notches
uhplc_calculus_long %>%
  filter(compound %in% c("caffeine", "cotinine", "nicotine", "salicyl", "theophyl")) %>% 
  ggplot(aes(x = pipe_notch, y = conc)) +
    geom_point() +
    facet_wrap(~ compound + batch, scales = "free_y")
```

@@

replicated_accuracy <-

# ratio of nicotine to cotinine
tobacco_ratio <- tobacco %>%
  select(id, batch, compound, conc) %>% 
  pivot_wider(names_from = compound, values_from = conc) %>% 
  group_by(id, batch) %>% 
  summarise(ratio = cotinine / nicotine) %>%
  remove_missing() %>% 
  left_join(select(demography, id, age, sex, preservation, pipe_notch)) %>% 
  filter(batch == "batch2",
         ratio != Inf) %>% 
  mutate(
    age = case_when(
      age == "eya" ~ 0,
      age == "lya" ~ 1,
      age == "ma" ~ 2,
      age == "old" ~ 3
    ),
    preservation = case_when(
      preservation == "fair" ~ 0,
      preservation == "good" ~ 1,
      preservation == "excellent" ~ 2
    )
  )
```

```{r}
#| label: fig-notch-ratio
#| fig-cap: "Relationship between ratio and number of pipe notches."

@@

In the replicated samples only, tobacco detection was successful in
`r nrow(filter(uhplc_accuracy, pipe_notch == "Y", presence > 0))`
out of
`r nrow(filter(uhplc_accuracy, pipe_notch == "Y"))`
pipe smokers
(`r scales::percent(nrow(filter(uhplc_accuracy, pipe_notch == "Y", presence > 0)) / nrow(filter(uhplc_accuracy, pipe_notch == "Y")), accuracy = 0.1)`)
Including individuals with absence of a pipe notch and concurrent absence of
compounds as a correct identification, gives an overall accuracy of
`r scales::percent(mean(uhplc_accuracy$correct, na.rm = T), accuracy = 0.1)`.

@@

Some of the teeth were missing because they have been sent elsewhere for DNA
sampling. These teeth were considered present when determining antemortem loss
ratios, and absent when scoring caries, periodontitis, and calculus.

```{r}
#| label: fig-dental-plots
#| layout-ncol: 2
#| fig-cap: 
#|   - "Antemortem tooth loss in sample."
#|   - "Periodontal scores in the sample."
dental_long %>%
  mutate(status = case_when(status == "dna" ~ "p")) %>% 
  dental_plot(fill = status)

@@

Ratio of antemortem lost teeth per present teeth at the site.

```{r}
#| label: fig-aml-id
#| fig-cap: "AMTL ratio per individual."
dental_long %>% 
  group_by(id) %>% 
  amtl_ratio(.status = status, .add = T)
```

```{r}
#| label: fig-aml-class
#| fig-cap: "AMTL ratio per tooth class."
dental_long %>% 
  group_by(class) %>% 
  amtl_ratio(.status = status, .add = T)
```

```{r}
#| label: fig-aml-type
#| fig-cap: "AMTL ratio per tooth type."
dental_long %>% 
  group_by(type) %>% 
  amtl_ratio(.status = status, .add = T)
```

@@

<!--add a caries correction?-->
Caries were scored as the location on each individual tooth. Multiple locations
on a single tooth were separated with `;`. The size of caries was also
recorded, but not used in further analysis. Large caries that cover multiple
surfaces with an unknown origin were recorded as 'crown'.

@@

```{r}
caries_count %>%
  caries_ratio(.caries = count, class, id) %>%
  ggplot(aes(x = class, y = ratio)) +
    geom_violin(aes(fill = class)) +
    geom_boxplot(width = 0.1) +
    theme_minimal()
```

```{r}
#| label: fig-caries-ratio
upper_order <- c(paste0("t", 18:11), paste0("t", 21:28))
lower_order <- c(paste0("t", 48:41), paste0("t", 31:38))

@@

Periodontitis was scored qualitatively on a scale from 0-3 as the amount of
horizontal bone loss from the CEJ to the alveolar bone, accounting for ca. 2mm
of gingival thickness.

@@

Pathological conditions and lesions that occur frequently in the population were
included in the analysis. Data were
dichotomised to presence/absence to allow statistical analysis. A conservative
approach was taken, so when in doubt, absence of a disease was assumed.
Osteoarthritis was considered present in cases where eburnation was visible
on one or more joint surfaces.
Vertebral osteophytosis is identified by marginal lipping and/or osteophyte
formation on the margin of the superior and inferior surfaces of the vertebral
body.
Cribra orbitalia was diagnosed based on the presence of pitting on the superior
surface of the orbit. No distinction was made between active or healing lesions.
Degenerative disc disease, or spondylosis, is identified as a large diffuse
depression of the
superior and/or inferior surfaces of the vertebral body [@rogersPalaeopathologyJoint2000].
Schmorl's nodes are identified as any cortical depressions on the surface of
the vertebral body. A note was made whether the lesion perforated the vertebral
margin, but both perforating and non-perforating lesions were recorded as present.
